Speed and strength of an epidemic intervention

### 2020-10-14 responses.tex
Hand back to Daniel

@@

JD: I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.

@@

We have expanded the application section with examples from COVID-19. To
our mind, these examples further underline the absence of r-based
thinking in the disease-modeling world. We also address more explicitly
the link between the parameters of the initial phase and the prospects
for long-term control. We feel that issues dealing directly with
non-exponential disease dynamics are complicated enough for another
paper.
We have expanded the
application section with examples from COVID-19, providing a qualitative
understanding on how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion on
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that R is constant, rather than r (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' The Lancet Global Health. In reality, given
observed r, more presymptomatic transmission reduces R, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of r-based
thinking in the disease-modeling world. While we appreciate the
reviewer's concern that a more data-driven approach could provide
further insight, we are worried that details may take away general ideas
behind speed- and strength-based paradigms. We feel that it is important
to provide a conceptual framework at this stage to a broader audience.
We also address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. We feel that
issues dealing directly with non-exponential disease dynamics are
complicated enough for another paper.

@@

SWP: I like this better for individual vs population-based: The efficacy
of individual-based intervention depends on time since infection (e.g.,
more likely to be isolated during symptomatic stage). The efficacy of
population-based intervention doesn't (e.g., the relative efficacy of
condom doesn't really change). What do you think?
JD: This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?

>>>>

\newcommand{\comment}{RENEW the comment command}
\renewcommand{\comment}[3]{}
\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

@@

\jd{I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.}

@@

We have expanded the application section with examples from COVID-19. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. We also address more explicitly
the link between the parameters of the initial phase and the prospects
for long-term control. We feel that issues dealing directly with
non-exponential disease dynamics are complicated enough for another
paper.
We have expanded the
application section with examples from COVID-19, providing a qualitative
understanding on how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion on
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that \RR\ is constant, rather than \rr\ (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' \emph{ Lancet Global Health}. In reality, given
observed \rr, more presymptomatic transmission reduces \RR, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. While we appreciate the
reviewer's concern that a more data-driven approach could provide
further insight, we are worried that details may take away general ideas
behind speed- and strength-based paradigms. We feel that it is important
to provide a conceptual framework at this stage to a broader audience.
We also address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. We feel that
issues dealing directly with non-exponential disease dynamics are
complicated enough for another paper.

@@

\swp{I like this better for individual vs population-based: The efficacy
of individual-based intervention depends on time since infection (e.g.,
more likely to be isolated during symptomatic stage). The efficacy of
population-based intervention doesn't (e.g., the relative efficacy of
condom doesn't really change). What do you think?}
\jd{This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?}

### 2020-10-14 speedstrength.tex
Hand back to Daniel

\newcommand{\comment}{RENEW the comment command}
\renewcommand{\comment}[3]{}
\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

@@

We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19. We showed that the strength-based paradigm provides clear insight into how this uncertainty affects \swp{non-targeted} interventions for the general population, while speed-based paradigm is a clearer way to think about interventions targeted to infected people.

While both strength- and speed-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, {\rr\ can be directly observed from case data but} the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).

@@

When comparing proposed interventions with estimated epidemic parameters to evaluate strategies, the situation is similar. Some scenarios lend themselves naturally to a single approach. For example, in the classic case of vaccination to eliminate a previously established childhood disease, both disease spread and intervention can be clearly characterized using strength \citep{anderson1985vaccination}. In our HIV example, both the HIV epidemic and the test-and-treat intervention can be best characterized using speed. Other cases, such as using social distancing (a strength-like intervention) in the early stages of COVID-19 (epidemic speed is observed) may not fit so neatly into either paradigm, however.
With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result
\swp{---the overall conclusion can change whether we calibrate our models to \rr or \RR.}
\swp{I feel like this is important but I'm not sure if this is a good place to put it. I think we need to say it explicitly.}

\swp{A new suggested paragraph. I think some version of this paragraph may help deal with above sentence (the image you sent me saying that you don't understand). I think it can also help deal with the reviewer's comments about the growth phase}
\swp{
Here, we focused on the initial growth phase an epidemic, assuming that the observed growth rate \rr is fixed throughout our analysis.
For endemic diseases, as discussed previously, strength-based framework can provide a clearer perspective---therefore, it makes more sense to rely on fixed \RR when evaluating the difficulty of elimination.
In these cases, parameter uncertainty will generally not affect the efficacy of strength-like interventions but can affect conclusions about speed-lik interventions:
given \RR, shorter generation intervals imply faster epidemic \citep{eaton2014proportion, powers2014impact, park2019practical}, requiring faster intervention to control .
While both strength- and speed-based frameworks will give the same answers for a given set of assumptions about an epidemic,
these conclusions can depend on whether we calibrate our models to the strength or speed of an epidemic.
}

There is an analogy here with measures of fitness in theoretical ecology.
For example, when a population is regulated by density dependence that affects all individuals identically, $r$ may be the best measure of fitness \citep{pasztor1996r0}, 
but when regulation primarily affects juvenile mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily}.
\swp{Do we need to explain more? feels a bit abrupt.}

The importance of speed-based perspectives are still rarely recognized in the case of infectious disease, however. 
Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak show an over-emphasis on strength at the expense of speed:   
during the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}. \swp{does this feel weak now? Maybe need something like ``Many subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing, with a few exceptions'' with citations?}
We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

>>>>

\newcommand{\comment}{RENEW the comment command}
\renewcommand{\comment}[3]{}
\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

@@

We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19. We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while the speed-based paradigm is a clearer way to think about interventions targeted to infected people. We concluded that increasing pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions. In hindsight, these conclusions are consistent with common sense, but in practice the speed-based framework provides a clear way to think about these questions.

While both strength- and speed-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, \rr\ can be directly observed from case data but the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).

Thinking explicitly about the two perspectives can also reduce confusion. Because of the dominance of the strength paradigm, researchers often explore different scenarios while holding \RR\ fixed. Fixed \RR\ is in fact a good default assumption for many endemic diseases. For invading diseases, however, \rr\ is likely to be better constrained by data than \RR. In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value) \citep{eaton2014proportion, powers2014impact, park2019practical}.

@@

When comparing proposed interventions with estimated epidemic parameters to evaluate strategies, the situation is similar. Some scenarios lend themselves naturally to a single approach. For example, in the classic case of vaccination to eliminate a previously established childhood disease, both disease spread and intervention can be clearly characterized using strength \citep{anderson1985vaccination}. In our HIV example, both the HIV epidemic and the test-and-treat intervention can be best characterized using speed. Other cases, such as using social distancing (a strength-like intervention) in the early stages of COVID-19 (epidemic speed is observed) may not fit so neatly into either paradigm, however.
With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result.

In population ecology, the duality between strength and speed is more widely recognized.
For example, when a population is regulated by density dependence that affects all individuals identically, $r$ may be the best measure of fitness, 
but when regulation primarily affects juvenile mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily,pasztor1996r0}.
There is also a link between the duality of these perspectives and the evolutionary tradeoff between speed and strength, commonly theorized as a tradeoff between \rr\ and carrying capacity $K$ \cite{Pianka70}.

The importance of speed-based perspectives are still rarely recognized in the case of infectious disease, however. 
Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak show an over-emphasis on strength at the expense of speed:   
during the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}. \swp{does this feel weak now? Maybe need something like ``Many subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing, with a few exceptions'' with citations?}
\jd{It would be great if you could do this.}
We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.
